---
title: ""
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Lab 02 - Loops and If-Else

## Loops

Many R functions operate on every element of a vector. For
example, to simulate from a log-normal distribution, you
can exponentiate draws from a normal. In the following code,
calling the function `exp` on `z` exponentiates each element of
`z`.

```{r}
z <- rnorm(10)
y <- exp(z)
y
```

These calculations can be replicated by writing a for loop over
the entries of `z`.

```{r}
print(z)
for( a in z ){
    print(c(a,exp(a)))
}
```
`for` is a function that takes its argument in a special syntax.
The object `z` is a vector (atomic or list), and `a` is a symbol
representing an element of the vector `z`. These two symbols are
separated by `in`, and then the parentheses are closed. Then we
open curly braces, include code involving `a`, and then close
the curly braces.

When the code is run, `a` is set to the first element of `z`, then
the code inside the curly braces is run, then `a` is set to the second
element of `z`, the code inside curly braces is run, then
`a` is set to the third element of `z`, and so on, until we get
to the last element of `z`.

Most often, programmers would replace `z` with a sequence of integers
and rewrite the loop as

```{r}
for( j in 1:length(z) ){
    print(c(z[j],exp(z[j])))
    y[j] <- exp(z[j])
}
```

Here, we loop over the integers `1:length(z)`. This allows us to do
useful things like assign to the jth element of `y`.

It's actually better to write the loop as

```{r}
for( j in seq_along(z) ){
    print(c(z[j],exp(z[j])))
    y[j] <- exp(z[j])
}

# show what seq_along does
seq_along(z)
seq_along( numeric(0) )
```

which does nothing when `length(z)` is zero, instead of the most likely
unintended behavior of looping over `1:0`.



# if

Very commonly, you will want your code to behave differently
based on the values of objects or elements of vectors. This conditional
behavior is often combined with loops.

Suppose we wanted to simulate from the half-normal distribution.
The half-normal pdf is simply the normal pdf for positive
values, and zero for negative. We can simulate from it by
taking the absolute value of draws from the normal:

```{r}
y <- rnorm(10)
print(y)

for( j in seq_along(y) ){

    # take negative of y[j] if y[j] < -
    if( y[j] < 0 ){
        y[j] <- -y[j]
    }

}
print(y)
```

`if` is a function takes takes a single logical value
as its argument. The statement `y[j] < 0` will evaluate
to `TRUE` or `FALSE` depending on the value of `y[j]`. If the
argment evaluates to `TRUE`, the code inside curly braces
is run. If `FALSE`, it's not run. And if the
statement evaluates to `NA`, an error is thrown

```{r, error=TRUE}
if(NA){
    print("This won't get printed")
}
```

If the argument is a vector whose length is longer than 1, a
warning is issued

```{r}
if( c(TRUE,FALSE) ){
    print("this works, but you better get your act together")
}
```

`if` will also do coercion of numeric to logical, and in both `for` and `if`,
you can omit braces if the code is only one line

```{r}
if(1){ print("this gets printed") }
if(2.7){ print("this gets printed") }
if(0){ print("this does not get printed") }
if(0.2){ print("this gets printed") }
if(1) print("this works too")
for(j in 1:10) print(j)
```

SIDENOTES: Absolute value can be done with the `abs` function. It can
also be done with subsetting:

```{r}
y <- rnorm(10)
print(y)
y[ y < 0 ] <- -y[ y < 0 ]
print(y)
```

# if-else

It's also common that you want the code to do something else
if the argument to the if statement evalutes to `FALSE`. Here
is a silly example

```{r}
speed_limit <- 65
speeds <- 65 + 5*rnorm(10)

for( j in seq_along(speeds) ){
    if( speeds[j] > speed_limit ){
        print( c(round(speeds[j]),"you're going too fast!") )
    } else {
        print( c(round(speeds[j]),"OK!") )
    }
}
```

This code checks whether than speed is greater than the limit. If it is
it lets the driver know; otherwise, it signals OK.

Ifs and elses can be strung together to evaluate a series of expressions.
Here is an implementation of the soft-threshold operator, thresholding
at -1 and 1:

```{r}
z <- rnorm(10)
y <- rep(NA, length(z))

for( j in seq_along(y) ){

    if( z[j] > 1 ){
        y[j] <- z[j] - 1
    } else if( z[j] < -1 ){
        y[j] <- z[j] + 1
    } else {
        y[j] <- 0
    }

}
print(cbind(z,y))
```

## Exercises

Use a for loop and if-else statements to implement the [trapezoidal
rule](https://en.wikipedia.org/wiki/Trapezoidal_rule) in order to approximately
integrate the function $x^2$ from 0 to 1 and the function $\sin(x)$
from 0 to $\pi$. Compare your answer to the actual value of the integral
for several choices of the number of integration ponts.
